---
title: "Report"
author: "Andrej Sokolič"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: false
    number_sections: true
    theme: spacelab
    fig_caption: true
    css: "report_style.css"
  pdf_document: default
params: 
  d: !r Sys.Date()
  eval: TRUE
  fig_align: center
  fig_asp: 0.618
  fig_out_width: 70%
  fig_pos: asis
  fig_width: 6
  locale: slovenian
  print_code: TRUE
  print_msgs: FALSE
  print_txt_results: TRUE
  print_session_info: TRUE
  print_warnings : FALSE
  subtitle: FALSE
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = params$print_code,
  eval = params$eval, 
  fig.align = params$fig_align,
  fig.asp = params$fig_asp,
  fig.show = params$fig_pos, 
  fig.width = params$fig_width,
  message = params$print_msgs,
  out.width = params$fig_out_width,
  results = params$print_txt_results, 
  warning = params$print_warnings
)

Sys.setlocale("LC_ALL", params$locale)

report_date <- ifelse(
  grepl("eng", params$locale, ignore.case = TRUE), 
  params$d, 
  format(params$d, format = "%d.%m.%Y")
)

report_subtitle <- ifelse(
  params$subtitle != FALSE,
  params$subtitle,
  ""
)
```


```{r libraries, echo = FALSE}
library(here)
library(data.table)
```

`r ifelse(params$print_session_info, "**Session Information:**\n", "")`

```{r sess_info, results='markup'}
if(params$print_session_info) sessioninfo::session_info()
```


---
date: `r report_date`
subtitle: `r report_subtitle`
---

[Article on medium](https://medium.com/analytics-vidhya/r-data-table-joins-48f00b46ce29)

# Assignment by reference

## The standard way

We create a data.table object and use the standard way to assign it to a new object.
We can see that the assignment by reference modifies the object that is still referenced by the original variable X. Changes in the new variable Y affect the original object.

```{r}
X <- data.table(
  name = c("A", "B", "C", "D"),
  value = c(1, 2, 3, 4)
)
X_ptr <- address(X)

Y <- X
Y_ptr <- address(Y)
X_ptr == Y_ptr

Y[name == "C", value := 0]
Y_ptr_new <- address(Y)
Y_ptr_new == Y_ptr
Y_ptr_new == X_ptr

Y[name == "C", value]
X[name == "C", value]
```

## Using copy

Using copy creates a new object (new address) and therefore does not change the original object.
Changes in the new variable Z do not affect the original object.

```{r}
X <- data.table(
  name = c("A", "B", "C", "D"),
  value = c(1, 2, 3, 4)
)
X_ptr_copy <- address(X)
X_ptr_copy == X_ptr

Z <- copy(X)
Z_ptr <- address(Z)

X_ptr_copy == Z_ptr
Z[name == "C", value := 0]
Z_ptr_new <- address(Z)
Z_ptr_new == Z_ptr

Z[name == "C", value]
X[name == "C", value]
```


# Using joins

## Left join

```{r}
X <- data.table(
  name = c("A", "B", "C", "D"),
  value = c("x1", "x2", "x3", "x4")
)
X_ptr <- address(X)

Y <- data.table(
  name = c("A", "C", "D", "D", "E"),
  value = c("y1", "y2", "y3", "y4", "y5")
)
Y_ptr <- address(Y)
```


## Full join 

```{r}
merge.data.table(X, Y, by = "name", all = TRUE)
```

## Right / Left outer join

All names in Y and matching names in X.

```{r}
X[Y, on = .(name)]
X; Y
# We have to assign the new object to a new variable
Z1 <- X[Y, on = .(name)]
Z1
Z1_ptr <- address(Z1)
Z1_ptr == X_ptr
Z1_ptr == Y_ptr

# Using merge.data.table
merge.data.table(X, Y, by = "name", all.y = TRUE)
```

All names in X and matching names in Y.


```{r}
Y[X, on = .(name)]
X; Y
# We have to assign the new object to a new variable
Z2 <- X[Y, on = .(name)]
Z2
Z2_ptr <- address(Z2)
Z2_ptr == X_ptr
Z2_ptr == Y_ptr

# Using merge.data.table
merge.data.table(X, Y, by = "name", all.x = TRUE) 
```

### Using mult argument

```{r}
# Default
Y[X, on = .(name), mult = "all"]
# First match
Y[X, on = .(name), mult = "first"]
# Last match
Y[X, on = .(name), mult = "last"]
```

## Inner join

```{r}
X[Y, on = .(name), nomatch = NULL]
merge.data.table(X, Y, by = "name")
```

## Anti join

Return rows of X that do not have a match in Y.

```{r}
X; Y
X[!Y, on = .(name)]
```

## Semi - join

Return rows of the left table that have a match in the right table. The which parameter returns matching rows with NA for non matching rows.

```{r}
X[na.omit(X[Y, on = .(name), which=TRUE])]

Y[na.omit(Y[X, on = .(name), which=TRUE])]
```


## Using keys

The join is automatically done on key columns.

```{r}
atools::check_error(X["A"])

setkey(X, name)
setkey(Y, name)

X["A"]
Y["D"]
X[Y]
Y[X, mult = "all"]
Y[X, mult = "first"]
```

### Full outer join

```{r}
unique_ids = unique(c(X[, name], Y[, name]))
X[Y[unique_ids]]
merge.data.table(X, Y, all = TRUE)
```


## Using non equi joins

```{r}
fenway <- fread(
    "Visitor, Start, End
    Cleveland, 2021-09-03, 2021-09-05
    Tampa Bay, 2021-09-06, 2021-09-08
    Baltimore, 2021-09-17, 2021-09-19
    NY Mets, 2021-09-21, 2021-09-22
    NY Yankees, 2021-09-24, 2021-09-26")

september <- 
   data.table(date = seq(as.Date('2021-09-01'), 
   as.Date('2021-09-30'), 'days'))
head(september)

home_games <- 
  fenway[september, on = .(Start <= date, End >= date), 
  .(Visitor, Date = Start)][!is.na(Visitor)]
home_games

# Columns from LHS must be on LHS
atools::check_error(fenway[september, on = .(date >= Start, date <= End), 
       .(Visitor, Date = Start)][!is.na(Visitor)])
```


### Rolling over

Carries forward the last match until new match.

```{r}
home_games[september, on = .(Date == date)]
home_games[september, on = .(Date == date), roll = TRUE]

# Opposite direction
home_games[september, on = .(Date == date), roll = -Inf]
```


### OVerlapping range join

```{r}
yankees <- fread(
  "NYY Opponent, Start, End
  Anaheim, 2021-09-01, 2021-09-01
  Baltimore, 2021-09-03, 2021-09-05
  Toronto, 2021-09-06, 2021-09-09"
)
setkey(yankees, Start, End)

fenway; 
yankees
foverlaps(fenway, yankees)

```



## Join by reference

```{r}
X[Y, on = .(name), Y_value := i.value]
X;Y
X_ptr == address(X)

# Here mult does nothing
X[Y, on = .(name), Y_value := i.value, mult = "first"]
X
X[Y, on = .(name), Y_value := i.value, mult = "last"]
X
```


```{r}
X; Y
Y[X, on = .(name), X_value := i.value]
X;Y
Y_ptr == address(Y)
```


### Warning

Notice below how the standard way of assigning variables still works on the original objects.

```{r}
X2 <- X
X2
X2[Y, on = .(name), Y_value2 := i.value]
X; X2
```




